The Revisited Hidden Weight Bit Function

### 2024-12-08 RevisedHiddenWeight.tex
Finalized proofreading

\begin{itemize}
	\item\emph{Balancedness analysis} (Section~\ref{sec:defAndBal}): We define the revisited \hwbf{} in $n$ variables, hereafter denoted by $f$, and analyze its balancedness. This involves studying the (restricted) Walsh transform of specific quadratic functions over sets of fixed Hamming weights (called slices). By establishing recursive relations for these values, we determine for which values of $n$ the function is balanced.
    \item \emph{Nonlinearity bounds} (Section~\ref{sec:dnk}):
    We relate the maximum Walsh coefficients of the revisited \hwbf{} to the coefficients of generating functions and employ complex analysis techniques to prove lower bounds on the nonlinearity of $f$. 
    Unlike most studies on Boolean functions used in cryptography, this approach yields strong nonlinearity bounds for a broader class of functions and for all even $n$.
    \item \emph{Experimental results and comparisons} (Section~\ref{sec:expAndComp}): 
    For bounded values of $n$, we employ specific techniques to refine the nonlinearity bounds of $f$, providing tighter estimates up to $n = 80$. Through experiments, we compare these bounds with actual nonlinearity values for $f$ and other weightwise quadratic functions, such as the majority function, the \hwbf{}, and the two main examples in~\cite{DAM:MeaOza24}. 
    Our results highlight the revisited \hwbf{}'s superior nonlinearity relative to functions with similar computational costs.
    \item \emph{Analysis of other cryptographic parameters} (Section~\ref{sec:otherParameters}):  
    We evaluate other cryptographically relevant parameters of $f$, including degree, algebraic immunity, and fast algebraic immunity. Similar to the nonlinearity analysis, the revisited \hwbf{} outperforms or matches other functions with comparable computational costs.
\end{itemize}

@@

Throughout this paper, $n\geq 0$ will always denote a non-negative integer. Further, our intervals will only contain integers, so that for instance $[0,n]=\{0,1,\dots,n\}$. The set of binary vectors of length $n$ will be denoted by $\F_2^n$, with zero vector written as $0_n$. The canonical basis of $\F_2^n$ will be written as $\{e_1,\dots,e_n\}$; therefore, $e_i\in\F_2^n$ is the vector which is zero everywhere, except at position $i$. The entries of a binary vector $x\in\F_2^n$ will always be denoted by $x_1,\dots,x_n$. Given a permutation $\pi:[1,n]\to[1,n]$ and a vector $x\in\F_2^n$, we define $\pi(x)=(x_{\pi(i)})_{i\in[1,n]}\in\F_2^n$. We write the scalar product of two binary vectors $x,y\in\F_2^n$ as $x\cdot y=\sum_{i=1}^nx_iy_i\in\F_2$. The Hamming weight of a binary vector $x\in\F_2^n$ is denoted by $\w(x)=|\{i\in [1,n]:x_i=1\}|$.

@@

In this part we recall general concepts on Boolean functions and their cryptographic properties we use in this article. 
For a deeper introduction on Boolean functions and their cryptographic parameters, we refer to the book~\cite{Carlet20}, and to~\cite{TOSC:CarMeaRot17} for properties on slices, so-called weightwise properties.

\begin{definition}[Slice]
    For $k \in [0,n]$, we introduce the set $\Ekn{k}{n}=\{x\in \F_2^n:\w(x)=k\}$, and call it the $k$-th slice of the Boolean hypercube (of dimension $n$). 
\end{definition}

\begin{definition}[Boolean function]\label{def:bool_f}
	A Boolean function in $n$ variables is a function from $\F_2^n$ to $\F_2$. The set of all Boolean functions in $n$ variables is denoted by $\BN$, and we denote this set without the null function by $\BN^*$.
\end{definition}

@@

\begin{definition}[Algebraic Normal Form, degree]\label{def:anf}
	We call Algebraic Normal Form of a Boolean function $f\in\BN$ its unique representation as an element of the ring $\F_2[x_1,\dots,x_n]/(x_1^2+x_1,\dots,x_n^2+x_n)$, and we represent it as follows:
	\[
		f(x_1,\dots,x_n)= \sum_{I \subseteq [1,n]} a_I \left( \prod_{i \in I} x_i \right),
	\]
	where $a_I\in \F_2$. The (algebraic) degree of $f$ is defined by $\degg(f)=0$ in case $f$ is the null function, and $\degg(f)=\max\{ |I|:I\subseteq[1,n],a_I=1\}$ otherwise.
\end{definition}

\begin{definition}[Walsh transform, restricted Walsh transform]\label{def:walsh_transform}
	The Walsh transform at $a \in \F_2^n$ of a Boolean function $f\in\BN$ restricted to a subset $S\subseteq \F_2^n$ is defined as:
    \[  \wt{f,S} (a) = \sum_{x\in S} (-1)^{f(x)+a \cdot x}.\]
    The Walsh transform of $f$ is then defined as $\wt{f}=\wt{f,\F_2^n}$. For $k\in[0,n]$, we also set $\wt{f,k}=\wt{f,\Ekn{k}{n}}$.
\end{definition}

@@

\begin{definition}[Nonlinearity, \eg{} page 79 in \cite{Carlet20}]%and weightwise nonlinearity
	\label{def:nl}
	The nonlinearity $\NL(f)$ of a Boolean function $f\in \BN$ is the minimum Hamming distance between $f$ and all the affine functions in $\BN$:
	\[
		\NL(f) = \min\{ \hd(f,g):g\in\BN,\degg(g)\leq 1\},
	\]
	where the Hamming distance between $f$ and $g$ is defined as $\hd(f,g)=|\{x\in\F_2^n:f(x)\neq g(x)\}|$. Alternatively, the nonlinearity of $f\in\BN$ can also be defined in terms of its Walsh transform:
	\[
		\NL(f) = 2^{n-1}- \frac{1}{2} \max_{a\in \F_2^n}{|\wt{f}(a)|}.
	\]
\end{definition}

When $n$ is even, the nonlinearity of a Boolean function in $n$ variables can reach at most $2^{n-1}-2^{n/2 -1}$. Functions reaching this maximum are called bent functions. Due to their broad applications and importance, they have been the focus of multiple works, \eg{} \cite{Rothaus76,FSE:Dobbertin95,DBLP:conf/itw/BudaghyanCH11,Tokareva15,Mesnager16}. 
The quadratic function $d_n\in\BN$ given by $d_n(x)=\sum_{i=1}^{n/2}x_{2i-1}x_{2i}$ is an example for such a bent function.

@@

The algebraic degree of multiples of a Boolean function is a crucial quantity for various attacks on stream ciphers that utilize a Boolean function as a filter. 
The well-known algebraic attack~\cite{EC:CouMei03} and fast algebraic attack~\cite{C:Courtois03} on filtered \textsf{LFSR}s have motivated the study of cryptographic properties such as algebraic immunity and fast algebraic immunity, investigated in numerous works, \eg{} \cite{FSE:Armknecht04,C:HawRos04,ISC:JiaZhaWan13})). Nowadays, these parameters are systematically determined for any function considered as a filter.

@@

\begin{definition}[Fast algebraic immunity, \eg{} \cite{EC:ACGKMR06}]\label{def:FAI}
	The fast algebraic immunity of a Boolean function $f\in \Bfn{n}$ is defined as:
	\[
		\FAI(f)=\min \{ 2 \AI(f), \min\{\degg(g)+ \degg(fg):g\in\BN,1\leq\degg(g)<\AI(f)\}\}.
	\]
\end{definition}

\subsection{Symmetric Functions, \hwbf{} and weightwise degree-d functions}

Recall that the Boolean symmetric functions in $n$ variables are those that are constant on the slice $\Ekn{k}{n}$ for every $k\in [0,n]$. 
This class of functions has been thoroughly studied in the context of cryptography, see \eg{} \cite{IEEE:Carlet04,IEEE:CanVid05,INDO:BraPre05,DM:SarMai07,IEEE:QFLW09,IEEE:CheLu11,Latin:Meaux19,CCDS:Meaux21,IEEE:CarMea21}.
In this article, the symmetric functions that we consider will mainly be the slice indicator function and the majority function.

@@

Various weightwise affine functions (\ie{} belonging to $\WD{1}{n}$) have been exhibited, such as in~\cite{TOSC:CarMeaRot17} where the bent functions in Propositions $1$ and $2$ are weightwise affine, or in~\cite{DAM:GinMea22} to show than no weightwise perfectly balanced function is weightwise affine for $n\geq 8$. 
The arguably best known example of weightwise affine function is the hidden weight bit function introduced in~\cite{IEEE:Bryant91}, the one obtained by fixing $f_0=0$ and $f_k=x_k$ for $k \in [1,n]$. The cryptographic properties of this function have been studied in~\cite{DAM:WCST14}, showing good algebraic properties for this function.

\begin{definition}[Hidden Weight Bit Function]\label{def:hwbf}
    We call Hidden Weight Bit Function (\hwbf{}) the Boolean function $h\in\BN$ defined by:
	\[h(x)= \left\{ 
	\begin{array}{l l}
		0 & \quad \text{ if } \w(x) =0, \\
		x_k & \quad \text{ if } \w(x)=k.
	\end{array} \right. \] 
\end{definition}

@@

\begin{Prop}\label{prop:wwd1}
	Let $n\geq 2$ be an even integer.
	\begin{itemize}
		\item (\eg{} Theorem $3$ in~\cite{DCC:DalMaiSar06}) The majority function $\mathsf{Maj}_n$ introduced in Definition~\ref{def:maj} satisfies:
		\[
			\NL(\mathsf{Maj}_n)=2^{n-1}-\binom{n-1}{\frac n2}.
		\]
		\item (Theorem $3$ in~\cite{DAM:WCST14}) The \hwbf{} $h$ introduced in Definition~\ref{def:hwbf} satisfies:
		\[
			\NL(h)=2^{n-1}-2\binom{n-2}{\frac{n-2}{2}}.
		\]	
		\item (Theorem $1$ in \cite{DAM:WCST14}) The \hwbf{} $h$ introduced in Definition~\ref{def:hwbf} satisfies $\degg(h)=n-1$ for all $n\geq 4$.
	\end{itemize}
\end{Prop}

@@

We use Krawtchouk polynomials and some of their properties to prove one of our main results. We give the necessary definition here and refer to \eg{}~\cite{book:MacSlo78} for more details.

\begin{definition}[Krawtchouk polynomials]\label{def:Kraw}
	The $n$-th Krawtchouk polynomial of degree $k\in[0,n]$ is given by:
    \[
        \kraw{k}{x}{n}=\sum_{j=0}^{k} (-1)^j \binom{x}{j} \binom{n-x}{k-j}.
    \]
    If $k<0$, we further define $\kraw{k}{x}{n}=0$. The Krawtchouk polynomials can alternatively be characterized as the coefficients in the generating function for $(1+z)^{n-x} (1-z)^x$ in the variable $z$:
    \[
        (1+z)^{n-x} (1-z)^x=\sum_{k\in\Z} \kraw{k}{x}{n} z^k.
    \]
\end{definition}

@@

\begin{definition}[Revisited \hwbf{}]\label{def:revHWBF}
	For an even integer $n\geq 0$, we call revisited \hwbf{} the Boolean function $f\in\BN$ defined as:
	\[
		f(x)=\sum_{i=1}^{n/2}(x_i+1) x_{i+\frac{n}{2}} + \sum_{k=1}^nx_k\phikn{k}{n}(x).
	\]
\end{definition}

@@

\begin{proposition}\label{prop:restrWT}
	Let $n\geq 0$ be an even integer, and let $f\in\BN$ be the revisited \hwbf{}. Consider any binary vector $a\in\F_2^n$, and define $c=a+e_k+\sum_{i=n/2+1}^n e_i$. Then the restricted Walsh transform of $f$ satisfies:
	\[
		\wtk{f}{k}(a)=\sum_{x \in \Ekn{k}{n}} (-1)^{\sum_{i=1}^{n/2} x_i x_{i+n/2}+ c \cdot x}.
	\]
\end{proposition}

@@

Accordingly, we can study the restricted Walsh transform of the revisited \hwbf{} by analyzing the one of the following Boolean function $d_n\in\BN$:

\begin{definition}
Let $n\geq 0$ be an even integer. Then we define the Boolean function $d_n\in\BN$ by $d_n(x)=\sum_{i=1}^{n/2}x_{2i-1}x_{2i}$. Further, for an integer $k$ and a binary vector $a\in\F_2^n$, we define $\Dnka{n}{k}{a}=0$ if $n\not\in[0,n]$, and otherwise:
\[
    \Dnka{n}{k}{a}=\sum_{x\in \Ekn{k}{n}} (-1)^{d_n(x) +a\cdot x}.
\]
\end{definition}

The $\Dnka{n}{k}{a}$ satisfy a recursive relation:

\begin{proposition}\label{prop:recursiveDnka}
Let $n\geq 0$ be an even integer. Then for all $k\in[0,n]$ and $a\in\F_2^n$, the following hold true:
	\begin{itemize}
		\item We have $\Dnka 00a=1$ and $\Dnka 0ka=0$ if $k\neq 0$.
		\item If $n\geq 2$, then for $b=(a_1,\dots,a_{n-2})\in\F_2^{n-2}$, we have:
		\[
			\Dnka{n}{k}{a}=\Dnka{n-2}{k}{b}+\left((-1)^{a_{n-1}} + (-1)^{a_{n}}\right)\Dnka{n-2}{k-1}{b} + (-1)^{1+a_{n-1}+a_n} \Dnka{n-2}{k-2}{b}.
		\]
	\end{itemize}
\end{proposition}

\begin{proof}
    We have that $\Dnka 00a=(-1)^0=0$ since $\Ekn 00=\{\epsilon\}$ for the empty vector $\epsilon\in\F_2^0$. That $\Dnka 0ka=0$ if $k\neq 0$ follows directly from the definition. Assume now that $n\geq 2$. By considering binary vectors $x\in\F_2^n$ as $x=(y,x_{n-1},x_n)$ for $y\in\F_2^{n-2}$, we get:
    \begin{align*}
		\Dnka{n}{k}{a}&=\sum_{x\in \Ekn{k}{n}} (-1)^{d_n(x) +a\cdot x}\\
		&=\sum_{y\in \Ekn{k}{n-2}} (-1)^{d_{n-2}(y) +b\cdot y}+
		\sum_{y\in \Ekn{k-1}{n-2}} (-1)^{d_{n-2}(y) +b\cdot y + a_{n-1}}\\
		&\qquad+\sum_{y\in \Ekn{k-1}{n-2}} (-1)^{d_{n-2}(y) +b\cdot y + a_{n}}
		+\sum_{y\in \Ekn{k-2}{n-2}} (-1)^{d_{n-2}(y) +b\cdot y +1 + a_{n-1}+ a_{n}}\\
		&=\Dnka{n-2}{k}{b}+\left((-1)^{a_{n-1}} + (-1)^{a_{n}}\right)\Dnka{n-2}{k-1}{b} + (-1)^{1+a_{n-1}+a_n} \Dnka{n-2}{k-2}{b}.\qedhere
    \end{align*}
\end{proof}

\begin{remark}
    We note that Proposition~\ref{prop:recursiveDnka} gives three different cases depending on the values of the two last elements of $a$:%Then it gives three cases:
	\begin{itemize}
		\item if $a_{n-1}=0=a_n$, then $\Dnka{n}{k}{a}=\Dnka{n}{k}{b}+2\Dnka{n}{k-1}{b}-\Dnka{n}{k-2}{b}$,
		\item if $a_{n-1} \ne a_n$, then $\Dnka{n}{k}{a}=\Dnka{n}{k}{b}+\Dnka{n}{k-2}{b}$,
		\item if $a_{n-1}=1=a_n$, then $\Dnka{n}{k}{a}=\Dnka{n}{k}{b}-2\Dnka{n}{k-1}{b}-\Dnka{n}{k-2}{b}$.
	\end{itemize}
\end{remark}

\begin{remark}\label{rmk:dependence_dnka}
    The recursive formula of Proposition~\ref{prop:recursiveDnka} gives different cases depending on the values of the pair $(a_{n-1},a_n)$. However, the same reasoning applies with any pair of the form $(a_{2i-1}, a_{2i})$ for $i\in[0,n/2]$. Therefore, the value of $\Dnka{n}{k}{a}$ depends only on the number of pairs $(a_{2i-1}, a_{2i})$ in $a$ being $(0,0)$, $(1,1)$, $(0,1)$ or $(1,0)$. We give the values of $\Dnka{2}{k}{a}$ for $n=2$ in Table~\ref{tab:Dnka}. These values together with Proposition~\ref{prop:recursiveDnka} are sufficient to determine any $\Dnka{n}{k}{a}$. We will get back to this in Remark~\ref{rmk:Dnka_dep_pqr}.
\end{remark}

\begin{table}
	\small
	\centering
	\begin{tabular}{|c|c|c|c|}
		\hline
		$a$ & $(0,0)$ & $(0,1)$ & $(1,1)$ \\ \hline
		$\Dnka{2}{0}{a}$ &$1$&$1$ & $1$\\
		$\Dnka{2}{1}{a}$ &$2$& $0$ & $-2$\\
		$\Dnka{2}{2}{a}$ &$-1$&$1$ & $-1$\\ \hline
	\end{tabular}
	\mycaption{Values of $\Dnka{2}{k}{a}$ for $n=2$.}\label{tab:Dnka}
\end{table}

In the following, we show how the Walsh transform of $f$ can be written in terms of the $\Dnka{n}{k}{a}$. Then, in Section~\ref{sec:balancedness} we use Proposition~\ref{prop:recursiveDnka} to determine the balancedness of $f$, and in Section~\ref{sec:genfunc} we study the value of $\Dnka{n}{k}{a}$ using generating functions.

\begin{proposition}\label{prop:WT}
	Let $n\geq 0$ be an even integer, and let $f\in\BN$ be the revisited \hwbf{}. Denote by $\pi:[1,n]\to[1,n]$ the permutation sending the first $n/2$ elements to the odd positions and the $n/2$ last ones to the even positions. Let $a\in\F_2^n$ be a binary vector, and let $b=\pi^{-1}(a)+(0,1,0,1,\ldots,0,1)$. Then the following holds true:
	\[ \wt{f}(a)=1+\sum_{k=1}^n \Dnka{n}{k}{b+\pi(e_k)}.\]
\end{proposition}

>>>>

\begin{itemize}
	\item\emph{Balancedness analysis} (Section~\ref{sec:defAndBal}): We define the revisited \hwbf{} in $n$ variables, hereafter denoted by $f$, and analyze its balancedness. This involves studying the (restricted) Walsh transform of specific quadratic functions over sets of fixed Hamming weights (called slices). By establishing recursive relations for these values, we determine for which values of $n$ the function is balanced.
    \item \emph{Nonlinearity bounds} (Section~\ref{sec:dnk}):
    We relate the maximum Walsh coefficients of the revisited \hwbf{} to the coefficients of generating functions and employ complex analysis techniques to prove lower bounds on the nonlinearity of $f$. 
    Unlike most studies on Boolean functions used in cryptography, this approach yields strong nonlinearity bounds for a broader class of functions in an even number of variables.
    \item \emph{Experimental results and comparisons} (Section~\ref{sec:expAndComp}): 
    For bounded values of $n$, we employ specific techniques to refine the nonlinearity bounds of $f$, providing tighter estimates up to $n = 80$. Through experiments, we compare these bounds with actual nonlinearity values for $f$ and other weightwise quadratic functions, such as the majority function, the \hwbf{}, and the two main examples in~\cite{DAM:MeaOza24}. 
    Our results highlight the revisited \hwbf{}'s superior nonlinearity relative to functions with similar computational costs.
    \item \emph{Analysis of other cryptographic parameters} (Section~\ref{sec:otherParameters}):  
    We evaluate other cryptographically relevant parameters of $f$, including degree, algebraic immunity, and fast algebraic immunity. Similar to the nonlinearity analysis, the revisited \hwbf{} outperforms or matches other functions with comparable computational costs.
\end{itemize}

@@

Throughout this paper, $n\geq 0$ will always denote a non-negative integer. Further, our intervals will only contain integers, so that for instance $[0,n]=\{0,1,\dots,n\}$. The set of binary vectors of length $n$ will be denoted by $\F_2^n$, with the zero vector written as $0_n$. The canonical basis of $\F_2^n$ will be written as $\{e_1,\dots,e_n\}$; therefore, $e_i\in\F_2^n$ is the vector which is zero everywhere except at position $i$. The entries of a binary vector $x\in\F_2^n$ will always be denoted by $x_1,\dots,x_n$. Given a permutation $\pi:[1,n]\to[1,n]$ and a vector $x\in\F_2^n$, we define $\pi(x)=(x_{\pi(i)})_{i\in[1,n]}\in\F_2^n$. We write the scalar product of two binary vectors $x,y\in\F_2^n$ as $x\cdot y=\sum_{i=1}^nx_iy_i\in\F_2$. The Hamming weight of a binary vector $x\in\F_2^n$ is denoted by $\w(x)=|\{i\in [1,n]:x_i=1\}|$.

@@

In this part, we recall general concepts on Boolean functions and their cryptographic properties we use in this article. 
For a deeper introduction on Boolean functions and their cryptographic parameters, we refer to the book~\cite{Carlet20}, and to~\cite{TOSC:CarMeaRot17} for properties on slices, so-called weightwise properties.

\begin{definition}[Slice]
    For any integer $k$, we introduce the set $\Ekn{k}{n}=\{x\in \F_2^n:\w(x)=k\}$, and call it the $k$-th slice of the Boolean hypercube (of dimension $n$); note that if $k\not\in[0,n]$, then $\Ekn kn=\varnothing$.
\end{definition}

\begin{definition}[Boolean function]\label{def:bool_f}
	A Boolean function in $n$ variables is a function from $\F_2^n$ to $\F_2$. The set of all Boolean functions in $n$ variables is denoted by $\BN$, and we denote this set without the null function by~$\BN^*$.
\end{definition}

@@

\begin{definition}[Algebraic Normal Form, degree]\label{def:anf}
	We call Algebraic Normal Form of a Boolean function $f\in\BN$ its unique representation as an element of the ring $\F_2[x_1,\dots,x_n]/(x_1^2+x_1,\dots,x_n^2+x_n)$, and we express it as follows:
	\[
		f(x_1,\dots,x_n)= \sum_{I \subseteq [1,n]} a_I \left( \prod_{i \in I} x_i \right),
	\]
	where $a_I\in \F_2$. The (algebraic) degree of $f$ is defined by $\degg(f)=0$ in case $f$ is the null function, and $\degg(f)=\max\{ |I|:I\subseteq[1,n],a_I=1\}$ otherwise.
\end{definition}

\begin{definition}[Walsh transform, restricted Walsh transform]\label{def:walsh_transform}
	The Walsh transform at $a \in \F_2^n$ of a Boolean function $f\in\BN$ restricted to a subset $S\subseteq \F_2^n$ is defined as:
    \[  \wt{f,S} (a) = \sum_{x\in S} (-1)^{f(x)+a \cdot x}.\]
    The (unrestricted) Walsh transform of $f$ is then defined as $\wt{f}=\wt{f,\F_2^n}$. For any integer $k$, we also set $\wt{f,k}=\wt{f,\Ekn{k}{n}}$.
\end{definition}

@@

\begin{definition}[Nonlinearity, \eg, page 79 in \cite{Carlet20}]%and weightwise nonlinearity
	\label{def:nl}
	The nonlinearity $\NL(f)$ of a Boolean function $f\in \BN$ is the minimum Hamming distance between $f$ and all the affine functions in $\BN$:
	\[
		\NL(f) = \min\{ \hd(f,g):g\in\BN,\degg(g)\leq 1\},
	\]
	where the Hamming distance between $f$ and $g$ is defined as $\hd(f,g)=|\{x\in\F_2^n:f(x)\neq g(x)\}|$. Alternatively, the nonlinearity of $f\in\BN$ can also be defined in terms of its Walsh transform:
	\[
		\NL(f) = 2^{n-1}- \frac{1}{2} \max_{a\in \F_2^n}{|\wt{f}(a)|}.
	\]
\end{definition}

When $n$ is even, the nonlinearity of a Boolean function in $n$ variables can reach at most $2^{n-1}-2^{n/2 -1}$. Functions reaching this maximum are called bent functions. Due to their broad applications and importance, they have been the focus of multiple works, \eg, \cite{Rothaus76,FSE:Dobbertin95,DBLP:conf/itw/BudaghyanCH11,Tokareva15,Mesnager16}. 
The quadratic function $d_n\in\BN$ given by $d_n(x)=\sum_{i=1}^{n/2}x_{2i-1}x_{2i}$ is an example for such a bent function.

@@

The algebraic degree of multiples of a Boolean function is a crucial quantity for various attacks on stream ciphers that utilize a Boolean function as a filter. 
The well-known algebraic attack~\cite{EC:CouMei03} and fast algebraic attack~\cite{C:Courtois03} on filtered \textsf{LFSR}s have motivated the study of cryptographic properties such as algebraic immunity and fast algebraic immunity, \eg, \cite{FSE:Armknecht04,C:HawRos04,ISC:JiaZhaWan13}. Nowadays, these parameters are systematically determined for any function considered as a filter.

@@

\begin{definition}[Fast algebraic immunity, \eg, \cite{EC:ACGKMR06}]\label{def:FAI}
	The fast algebraic immunity of a Boolean function $f\in \Bfn{n}$ is defined as:
	\[
		\FAI(f)=\min \{ 2 \AI(f), \min\{\degg(g)+ \degg(fg):g\in\BN,1\leq\degg(g)<\AI(f)\}\}.
	\]
\end{definition}

\subsection{Symmetric Functions, \hwbf{} and weightwise degree-$d$ functions}

Recall that the Boolean symmetric functions in $n$ variables are those that are constant on the slice $\Ekn{k}{n}$ for every $k\in [0,n]$. 
This class of functions has been thoroughly studied in the context of cryptography, see \eg, \cite{IEEE:Carlet04,IEEE:CanVid05,INDO:BraPre05,DM:SarMai07,IEEE:QFLW09,IEEE:CheLu11,Latin:Meaux19,CCDS:Meaux21,IEEE:CarMea21}.
In this article, the symmetric functions that we consider will mainly be the slice indicator function and the majority function.

@@

Various weightwise affine functions (\ie, belonging to $\WD{1}{n}$) have been exhibited, such as in~\cite{TOSC:CarMeaRot17} where the bent functions in Propositions $1$ and $2$ are weightwise affine, or in~\cite{DAM:GinMea22} to show than no weightwise perfectly balanced function is weightwise affine for $n\geq 8$. 
The arguably best known example of weightwise affine function is the Hidden Weight Bit Function introduced in~\cite{IEEE:Bryant91}, the one obtained by fixing $f_0=0$ and $f_k=x_k$ for $k \in [1,n]$. The cryptographic properties of this function have been studied in~\cite{DAM:WCST14}, showing good algebraic properties for this function.

\begin{definition}[Hidden Weight Bit Function]\label{def:hwbf}
    We call Hidden Weight Bit Function (\hwbf{}) the Boolean function $h\in\BN$ defined as:
	\[
		h(x)=\sum_{k=1}^{n}x_k\phikn{k}{n}(x).
	\]
\end{definition}

@@

\begin{Prop}\label{prop:wwd1}
	Let $n\geq 2$ be an even integer.
	\begin{itemize}
		\item (\eg, Theorem $3$ in~\cite{DCC:DalMaiSar06}) The majority function $\mathsf{Maj}_n$ introduced in Definition~\ref{def:maj} satisfies:
		\[
			\NL(\mathsf{Maj}_n)=2^{n-1}-\binom{n-1}{\frac n2}.
		\]
		\item (Theorem $3$ in~\cite{DAM:WCST14}) The \hwbf{} $h$ introduced in Definition~\ref{def:hwbf} satisfies:
		\[
			\NL(h)=2^{n-1}-2\binom{n-2}{\frac{n-2}{2}}.
		\]	
		\item (Theorem $1$ in \cite{DAM:WCST14}) The \hwbf{} $h$ introduced in Definition~\ref{def:hwbf} satisfies $\degg(h)=n-1$ if $n\geq 4$.
	\end{itemize}
\end{Prop}

@@

We use Krawtchouk polynomials and some of their properties to prove one of our main results. We give the necessary definition here and refer to \cite{book:MacSlo78} for more details, for instance.

\begin{definition}[Krawtchouk polynomials]\label{def:Kraw}
	The $n$-th Krawtchouk polynomial of degree $k\in\Z$ is given by:
    \[
        \kraw{k}{x}{n}=\sum_{j=0}^{k} (-1)^j \binom{x}{j} \binom{n-x}{k-j}.
    \]
    The Krawtchouk polynomials can alternatively be characterized as the coefficients in the generating function for $(1+z)^{n-x} (1-z)^x$ in the variable $z$:
    \[
        (1+z)^{n-x} (1-z)^x=\sum_{k\in\Z} \kraw{k}{x}{n} z^k.
    \]
\end{definition}

@@

\begin{definition}[Revisited \hwbf{}]\label{def:revHWBF}
	For an even integer $n\geq 0$, we call revisited \hwbf{} the Boolean function $f\in\BN$ defined as:
	\[
		f(x)=\sum_{k=1}^nx_k\phikn{k}{n}(x)+\sum_{i=1}^{n/2}(x_i+1) x_{i+n/2}.
	\]
\end{definition}

@@

\begin{proposition}\label{prop:restrWT}
	Let $n\geq 0$ be an even integer, and let $f\in\BN$ be the revisited \hwbf{}. Consider any binary vector $a\in\F_2^n$ and any integer $k\in[1,n]$. Define $c=a+e_k+\sum_{i=n/2+1}^n e_i$. Then the restricted Walsh transform of $f$ satisfies:
	\[
		\wtk{f}{k}(a)=\sum_{x \in \Ekn{k}{n}} (-1)^{\sum_{i=1}^{n/2} x_i x_{i+n/2}+ c \cdot x}.
	\]
\end{proposition}

@@

Accordingly, we can study the restricted Walsh transform of the revisited \hwbf{} by analyzing the one of the following Boolean function $d_n\in\BN$.

\begin{definition}
	Let $n\geq 0$ be an even integer. Then we define the Boolean function $d_n\in\BN$ by:
	\[
		d_n(x)=\sum_{i=1}^{n/2}x_{2i-1}x_{2i}.
	\]
	Further, for an integer $k$ and a binary vector $a\in\F_2^n$, we define:
	\[
		\Dnka{n}{k}{a}=\wtk{d_n}{k}(a)=\sum_{x\in \Ekn{k}{n}} (-1)^{d_n(x) +a\cdot x}.
	\]
\end{definition}

The $\Dnka{n}{k}{a}$ satisfy a recursive relation.

\begin{proposition}\label{prop:recursiveDnka}
Let $n\geq 0$ be an even integer. Then for all integers $k$ and all binary vectors $a\in\F_2^n$, the following hold true:
	\begin{itemize}
		\item We have $\Dnka 00a=1$ and $\Dnka 0ka=0$ if $k\neq 0$.
		\item If $n\geq 2$, then for $b=(a_1,\dots,a_{n-2})\in\F_2^{n-2}$, we have:
		\[
			\Dnka{n}{k}{a}=\Dnka{n-2}{k}{b}+\left((-1)^{a_{n-1}} + (-1)^{a_{n}}\right)\Dnka{n-2}{k-1}{b} + (-1)^{1+a_{n-1}+a_n} \Dnka{n-2}{k-2}{b}.
		\]
	\end{itemize}
\end{proposition}

\begin{proof}
    We have that $\Dnka 00a=(-1)^0=1$ since $\Ekn 00=\{\varepsilon\}$ for the empty vector $\varepsilon=a\in\F_2^0$. That $\Dnka 0ka=0$ if $k\neq 0$ follows from $\Ekn k0$ being empty if $k\neq 0$. Assume now that $n\geq 2$. By considering binary vectors $x\in\F_2^n$ as $x=(y,x_{n-1},x_n)$ for $y\in\F_2^{n-2}$, we get:
    \begin{align*}
		\Dnka{n}{k}{a}&=\sum_{x\in \Ekn{k}{n}} (-1)^{d_n(x) +a\cdot x}\\
		&=\sum_{y\in \Ekn{k}{n-2}} (-1)^{d_{n-2}(y) +b\cdot y}+
		\sum_{y\in \Ekn{k-1}{n-2}} (-1)^{d_{n-2}(y) +b\cdot y + a_{n-1}}\\
		&\qquad+\sum_{y\in \Ekn{k-1}{n-2}} (-1)^{d_{n-2}(y) +b\cdot y + a_{n}}
		+\sum_{y\in \Ekn{k-2}{n-2}} (-1)^{d_{n-2}(y) +b\cdot y +1 + a_{n-1}+ a_{n}}\\
		&=\Dnka{n-2}{k}{b}+\left((-1)^{a_{n-1}} + (-1)^{a_{n}}\right)\Dnka{n-2}{k-1}{b} + (-1)^{1+a_{n-1}+a_n} \Dnka{n-2}{k-2}{b}.\qedhere
    \end{align*}
\end{proof}

\begin{remark}
    We note that Proposition~\ref{prop:recursiveDnka} gives three different cases depending on the values of the two last elements of $a$:%Then it gives three cases:
	\begin{itemize}
		\item if $a_{n-1}=0=a_n$, then $\Dnka{n}{k}{a}=\Dnka{n-2}{k}{b}+2\Dnka{n-2}{k-1}{b}-\Dnka{n-2}{k-2}{b}$,
		\item if $a_{n-1} \ne a_n$, then $\Dnka{n}{k}{a}=\Dnka{n-2}{k}{b}+\Dnka{n-2}{k-2}{b}$,
		\item if $a_{n-1}=1=a_n$, then $\Dnka{n}{k}{a}=\Dnka{n-2}{k}{b}-2\Dnka{n-2}{k-1}{b}-\Dnka{n-2}{k-2}{b}$.
	\end{itemize}
\end{remark}

\begin{remark}\label{rmk:dependence_dnka}
    The recursive formula of Proposition~\ref{prop:recursiveDnka} gives different cases depending on the values of the pair $(a_{n-1},a_n)$. However, the same reasoning applies with any pair of the form $(a_{2i-1}, a_{2i})$ for $i\in[1,n/2]$. Therefore, the value of $\Dnka{n}{k}{a}$ depends only on the number of pairs $(a_{2i-1}, a_{2i})$ in $a$ being $(0,0)$, $(1,1)$, $(0,1)$ or $(1,0)$. We give the values of $\Dnka{2}{k}{a}$ for $n=2$ in Table~\ref{tab:Dnka}. These values together with Proposition~\ref{prop:recursiveDnka} are sufficient to determine any $\Dnka{n}{k}{a}$.
\end{remark}

\begin{table}[ht]
	\small
	\centering
	\begin{tabular}{|c|c|c|c|}
		\hline
		$a$ & $(0,0)$ & $(0,1)$ & $(1,1)$ \\ \hline
		$\Dnka{2}{0}{a}$ &$1$&$1$ & $1$\\
		$\Dnka{2}{1}{a}$ &$2$& $0$ & $-2$\\
		$\Dnka{2}{2}{a}$ &$-1$&$1$ & $-1$\\ \hline
	\end{tabular}
	\mycaption{Values of $\Dnka{2}{k}{a}$.}\label{tab:Dnka}
\end{table}

In the following, we show how the Walsh transform of the revisited \hwbf{} $f\in\BN$ can be written in terms of the $\Dnka{n}{k}{a}$. Then, in Section~\ref{sec:balancedness}, we use Proposition~\ref{prop:recursiveDnka} to determine the balancedness of $f$, and in Section~\ref{sec:genfunc} we study the value of $\Dnka{n}{k}{a}$ using generating functions.

\begin{proposition}\label{prop:WT}
	Let $n\geq 0$ be an even integer, and let $f\in\BN$ be the revisited \hwbf{}. Denote by $\pi:[1,n]\to[1,n]$ the permutation sending the first $n/2$ elements to the odd positions and the $n/2$ last ones to the even positions. Let $a\in\F_2^n$ be a binary vector, and let $b=\pi^{-1}(a)+\sum_{i=1}^{n/2}e_{2i}$. Then the following holds true:
	\[ \wt{f}(a)=1+\sum_{k=1}^n \Dnka{n}{k}{b+\pi(e_k)}.\]
\end{proposition}
